\documentclass[11pt,a4paper]{article}

\usepackage[top=2cm,bottom=2cm,left=2.5cm,right=2.5cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{tikz}
\usepackage{titlesec}
\pgfplotsset{compat=1.18}
\usetikzlibrary{patterns,calc,arrows.meta}

% Part styling
\titleformat{\part}[display]
  {\centering\normalfont\Huge\bfseries}{\partname~\thepart}{20pt}{\Huge}
\titlespacing*{\part}{0pt}{40pt}{30pt}

% Colors
\definecolor{ourblue}{RGB}{70,130,180}
\definecolor{oborange}{RGB}{255,140,0}
\definecolor{asmblue}{RGB}{25,80,200}
\definecolor{taskred}{RGB}{200,50,50}
\definecolor{task2green}{RGB}{40,160,40}
\definecolor{sagreen}{RGB}{50,180,50}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  breaklines=true,
  frame=single,
  language=C,
  columns=flexible,
  keepspaces=true
}

% Diff 1: Title from File 2
\title{\textbf{AI-Driven HPC Development: A High-Performance BLAS Library}\\[6pt]
  \large{TER (Supervised Research Project)\\ M1 Quantum and Distributed Computer Science\\ Universit\'{e} Paris-Saclay}}
\author{Edna Iusupova \and Rosemarie Streefkerk}
\date{May 2026}

\begin{document}
\maketitle
\tableofcontents
\newpage

%%============================================================
\section*{Introduction}
\addcontentsline{toc}{section}{Introduction}
%%============================================================

This report documents the implementation and performance evaluation of a
single-precision BLAS (Basic Linear Algebra Subprograms) library developed over six weeks using AI-assisted development. The project is a case-study in AI-driven HPC development.
The entire codebase, testing infrastructure and autotuning loops were built iteratively through prompting with the Antigravity IDE
(using Claude Sonnet~4.6 and Gemini Pro). The objective was not only reaching hardware peak
performance but also testing how far we can get using AI-driven development and where human intervention is still needed to reach our goals.
All implementations are benchmarked against OpenBLAS to assess where AI-generated
kernels are competitive, where they fall short, and why.

\subsection*{Scope of Development}
The project spans six weeks of systematic, iterative development, categorized as follows:
\begin{itemize}
  \item \textbf{Part~1: GEMM and BLAS~3.} The core focus is SGEMM (Single Precision General Matrix Multiply), implemented via the BLIS (BLAS-like Library Instantiation Software) 5-loop architecture. This part also covers autotuning and derived BLAS~3 operations (\texttt{ssyrk, strsm, strmm}).
  \item \textbf{Part~2: BLAS~1.} Implementation of foundational vector-vector operations (\texttt{sscal, scopy, sswap, saxpy, sdot, snrm2, sasum, isamax, srot}), pushing single-thread and multi-thread memory bandwidth limits.
  \item \textbf{Part~3: BLAS~2.} Implementation of matrix-vector operations (\texttt{sgemv, ssymv, strmv, strsv, sger, ssyr, ssyr2}).
  \item \textbf{Part~4: Testing, autotuning and agent loop implementation.} Numerical validation testbench, autonomous tuning to increase performance based on parameters, and implementation of an agent loop to recreate the entire project from scratch starting from just an implementation plan. 
\end{itemize}

\subsection*{Hardware Architecture and Theoretical Limits}
Understanding the target hardware is critical for evaluating HPC software. The testbed is a KVM virtual machine with 20 vCPUs mapped to an Intel Xeon E5-2670 v3 (Haswell-EP) processor running at a fixed 2.30\,GHz clock (x86-64 architecture).

\paragraph{Compute Capabilities (AVX2 \& FMA).}
The processor supports Advanced Vector Extensions 2 (AVX2), featuring 256-bit wide vector registers. A single 256-bit register holds 8 single-precision floating-point numbers. The processor also supports Fused Multiply-Add (FMA) instructions, which perform $d = a \cdot b + c$ in a single hardware step, counting as 2 floating-point operations (FLOPs). 
A single core can issue two FMA instructions per clock cycle. The processor runs at a fixed base frequency of 2.30\,GHz; no turbo boost is available inside the KVM environment. The theoretical peak performance per core is:
\begin{equation}
\begin{split}
\text{Peak}_{\text{core}} &= 2\ (\text{instr./cycle}) \times 8\ (\text{elems./vec.}) \times 2\ (\text{FLOPs/FMA})\\
&\quad\times 2.30\times10^{9}\ (\text{cycles/s})\\
&= 73.6\ \text{GFLOP/s}
\end{split}
\end{equation}
Fully utilized across all 20 cores, the machine theoretical peak is $1472 \text{ GFLOP/s}$ (1.47 TFLOP/s).

\paragraph{Memory Hierarchy and Bandwidth.}
The memory hierarchy determines the bandwidth ceiling for BLAS~1 and BLAS~2 kernels. The CPU features three levels of cache:
\begin{itemize}
    \item \textbf{L1 Data Cache}: 32\,KB per core, 4--5 cycle load-to-use latency.
    \item \textbf{L2 Cache}: 256\,KB per core, $\sim$12--14 cycle latency.
    \item \textbf{L3 Cache}: $\approx$25\,MB, shared across all cores.
\end{itemize}
The main memory (RAM) utilizes DDR4-2133 modules. The Intel Xeon E5-2670 v3 (Haswell-EP) supports four DDR4 channels per socket, giving a theoretical maximum of $4 \times 17.1 = 68.3\,\text{GB/s}$ per socket. Effective sustained bandwidth, measured with the STREAM triad benchmark, reaches $\approx 40\,\text{GB/s}$ at 20 threads — the practical ceiling set by KVM memory-controller overhead and first-touch NUMA policy.
The 25\,MB L3 cache, shared across all 20 cores, has a substantially higher aggregate internal bandwidth: when all cores access the L3 simultaneously, the internal ring/mesh interconnect can sustain $\approx 300$--$500$\,GB/s. This creates two distinct bandwidth regimes that appear throughout this report:
\begin{itemize}
    \item \textbf{DRAM-bound}: data exceeds L3 capacity; effective bandwidth $\approx 40$\,GB/s (STREAM triad, 20T).
    \item \textbf{L3-bound}: data fits in L3 ($< 25$\,MB, e.g.\ vectors of $N = 2^{20}$ floats $= 4$\,MB each); effective bandwidth $\approx 300$--$500$\,GB/s aggregate.
\end{itemize}
Any algorithm that cannot reuse data in L1/L2 will hit either the $\approx 40$\,GB/s DRAM ceiling (STREAM triad, 20T) or the softer per-core L3 bandwidth ceiling, bottlenecking BLAS~1 and BLAS~2 kernels.

\paragraph{Software Environment.}
\begin{itemize}
  \item \textbf{Compiler}: GCC~13.3, utilizing aggressive optimization flags: \texttt{-O3 -march=native -mavx2 -mfma -fopenmp}.
  \item \textbf{Reference}: OpenBLAS (system package version), executed with identical \texttt{OMP\_NUM\_THREADS} to ensure a fair comparison.
  \item \textbf{Timing Methodology}: To combat OS jitter and CPU frequency scaling noise, all benchmarks employ an adaptive timing loop. A kernel is executed repeatedly until the coefficient of variation (CV) of the runtimes falls below 2\%, and the median time is reported.
\end{itemize}

\paragraph{Measured hardware limits and peak utilisation.}
Table~\ref{tab:hw_peaks} consolidates the theoretical ceilings and the best
measured results for each key hardware resource.
DRAM bandwidth was measured with a STREAM-triad loop;
SGEMM peaks are from the final 6$\times$16 TASK1 (2D parallelization) kernel on $1024^3$;
L3 aggregate bandwidth is inferred from \texttt{sger} at $N=2048$, 16T
(matrix occupies $\sim$16\,MB, fitting fully in L3).

% Diff 2: Table 1 from File 1
\begin{table}[h]
\centering
\caption{Server theoretical limits and best measured effective peaks (single precision).}
\label{tab:hw_peaks}
\small
\begin{tabular}{lrrrl}
\toprule
Resource & Theoretical limit & Measured peak & \% of limit & Condition \\
\midrule
FP32 throughput, 1 core  & 73.6\,GF/s   & 69.7\,GF/s  & 95\% & SGEMM $1024^3$, 1T \\
FP32 throughput, 8 cores & 588.8\,GF/s  & 458.3\,GF/s & 78\% & SGEMM $1024^3$, 8T \\
DRAM bandwidth (STREAM)  & 68.3\,GB/s   & 40.1\,GB/s  & 59\% & STREAM triad, 8T \\
DRAM bandwidth (STREAM)  & 68.3\,GB/s   &  7.8\,GB/s  & 11\% & STREAM triad, 1T \\
L3 aggregate bandwidth   & ---           & 491\,GB/s   & ---  & \texttt{sger} $2048^2$, 16T \\
\bottomrule
\multicolumn{5}{l}{DRAM theoretical: 4-channel DDR4-2133, $4\times17.1$\,GB/s per socket.}\\
\multicolumn{5}{l}{SGEMM benchmarked at 1T, 4T, 8T; see Table~\ref{tab:week3}.}
\end{tabular}
\end{table}

\newpage

%%============================================================
\part{GEMM and BLAS~3}
%%============================================================

\section{SGEMM development}

\subsection{BLIS framework}

Matrix multiplication $C = \alpha A B + \beta C$ requires $O(N^3)$ arithmetic but only $O(N^2)$ data. With proper tiling, the arithmetic intensity (FLOPs per byte) is high enough that the AVX2 FMA units, not memory, become the bottleneck.

% Diff 3: Notation paragraph from File 1
\paragraph{Notation.}
Throughout Part~1, $M$, $K$, $N$ denote the dimensions of the operands:
$A$ is $M\times K$, $B$ is $K\times N$, $C$ is $M\times N$.
The BLIS tiling parameters are:
$MC$, $KC$, $NC$ — macro-block sizes along $M$, $K$, $N$, chosen so that
the corresponding data panels fit in a target cache level;
$MR$, $NR$ — micro-kernel tile dimensions, chosen so that the
$MR\times NR$ accumulator block fits entirely in the AVX2 register file.

We adopted the BLIS architecture, which decomposes the GEMM operation into five nested loops surrounding a hardware-specific micro-kernel. This hierarchical decomposition ensures that each level of the memory hierarchy (L3, L2, and registers) is fully exploited — the packed buffers eliminate redundant cache misses, and the micro-kernel operates entirely on data that fits in registers, maximizing arithmetic intensity and minimizing memory traffic:
\begin{enumerate}
    \item \textbf{Loop 5 (JC)}: Slices $N$ into blocks of size $NC$.
    \item \textbf{Loop 4 (PC)}: Slices $K$ into blocks of size $KC$. Matrix $B$ is packed into a contiguous L3-resident buffer.
    \item \textbf{Loop 3 (IC)}: Slices $M$ into blocks of size $MC$. Matrix $A$ is packed into a contiguous L2-resident buffer.
    \item \textbf{Loop 2 (JR)}: Steps through the $NC$ block in chunks of $NR$.
    \item \textbf{Loop 1 (IR)}: Steps through the $MC$ block in chunks of $MR$.
\end{enumerate}
At the very center lies the Micro-Kernel, computing an $MR \times NR$ block of $C$ by taking the dot product of an $MR \times KC$ panel of $A$ and a $KC \times NR$ panel of $B$.

\subsection{8$\times$8 AVX2 micro-kernel}

We provided the AI agent with the BLIS algorithmic blueprint and tasked it to generate the baseline SGEMM. 
It \textbf{succeeded} on first try in producing a functionally correct 5-loop framework. 

\paragraph{Micro-kernel Register blocking.}
To compute an $8 \times 8$ block of $C$ (to fit in the micro-kernel), we need to store 64 single-precision floats. Since one 256-bit (YMM) register holds 8 floats, the $C$ accumulator block requires $64 / 8 = 8$ YMM registers.
During the inner $K$-loop, we must load a column of $A$ (1 register) and broadcast elements of $B$ (1 register). This requires $8 \text{ (accumulators)} + 1 \text{ (A)} + 1 \text{ (B)} = 10$ registers, fitting comfortably within the 16 YMM registers available in AVX2.

However, with only 10 of the 16 YMM registers in use, 6 registers remain idle, reducing the number of independent FMA accumulator chains available to hide the processor's FMA latency and limiting instruction-level parallelism. At this point in the project, single-thread performance on a $1024^3$ matrix reached $\sim$40\,GF/s (54\% of the 73.6\,GF/s core peak). While impressive for a first iteration, it fell short of OpenBLAS's $\sim$50\,GF/s, due to register underutilization.

\subsection{Micro-kernels and parallelism}

\paragraph{Register-bound micro-kernels (6$\times$16 and 4$\times$24).}
To improve the ratio of FMAs to memory loads, we instructed the AI to maximise use of the 16 YMM registers. Two kernels were produced:
\begin{itemize}
  \item \textbf{6$\times$16 Micro-kernel}: We cannot divide $M=6$ evenly into 8-float registers but we can treat $M=6$ as 1 full register with 2 floats ignored. We then obtain $MR=6$ output \emph{rows}, $NR=16$ output columns. For each $k$-step, each of the 6 rows of $A$ contributes a \emph{scalar broadcast} (one YMM register, reused across the $NR/8=2$ FMA chains); $B$ is loaded as $NR/8 = 2$ YMM vectors. The output accumulator block requires $MR \times (NR/8) = 6 \times 2 = 12$ YMM registers. The total register pressure is: $12 \text{ (C accumulators)} + 2 \text{ (B)} + 1 \text{ (A broadcast)} = 15$ YMM, leaving 1 free. Each $k$-step performs $6 \times 2 = 12$ FMAs.
  \item \textbf{4$\times$24 Micro-kernel}: $MR=4$ output rows (half a register), $NR=24$ columns. Each of the 4 rows of $A$ is a scalar broadcast; $B$ requires $NR/8 = 3$ YMM loads. Accumulator block: $4 \times 3 = 12$ registers. Total: $12 \text{ (C)} + 3 \text{ (B)} + 1 \text{ (A broadcast)} = 16$ YMM registers, perfectly saturating the AVX2 register file with zero register spills.
\end{itemize}
Both configurations were generated \textbf{correctly on the first try}, complete with pointer arithmetic and inline-assembly equivalents. However, the AI agent needed an explicit prompt to come up with the two kernels.

% Diff 4: TASK1 description from File 2
\paragraph{2D task parallelism (referred to as TASK1 in our project).}
The output matrix C is a 2D grid of tiles (M $\times$ N). The idea of 2D parallelism is to distribute those tiles across threads. Each thread picks up whichever tile is available next, computes its full dot product and writes the result directly to C. There is no reduction step. Initially, the AI agent used \texttt{omp taskloop collapse(2)}, which introduced 10$\times$ overhead (which was caught by benchmarking). On a new try it was replaced by \texttt{omp parallel for collapse(2) schedule(dynamic)}.

\paragraph{K replication (referred to as TASK2 in our project).}
The K dimension is split into $r$ slices, each writing a private partial-$C$
buffer, then merged via a binary reduction tree. The structure was generated
incorrectly by AI. On the first try it generated another implementation of 2D parallelization instead of the 3D parallelization we prompted for. Upon further prompting, it introduced the k dimension but created a new issue by  using a \emph{sequential} reduction step instead of a binary tree, thus loosing all the gains of parallelization. Even after asking the agent to analyze its own results, it did not notice this issue and it was only caught later by human review of the code. After this we explicitly prompted the AI to use a binary reduction tree and this time the issue was fully solved and K replication was implemented successfully.

\paragraph{20-thread benchmark.}
Table~\ref{tab:week2} shows the first complete benchmark run in the second week of the project. The TASK2 column
reveals the issue with sequential reduction.

% Diff 5: Caption from File 2, sentence after table from File 1
\begin{table}[h]
\centering
\caption{SGEMM (20 threads, 7 runs/median). OB = OpenBLAS.}
\label{tab:week2}
\begin{tabular}{lrrrrrr}
\toprule
Size & OB & 8$\times$8 & 6$\times$16 & 4$\times$24 & ASM & TASK2\\
\midrule
$64^3$   & 50  & 29  & 34  & 30  & 33  & 34\\
$128^3$  & 77  & 85  & 79  & 64  & 77  & 0.07\dag\\
$256^3$  & 151 & 251 & 274 & 274 & 271 & 0.40\dag\\
$512^3$  & 360 & 429 & 568 & 512 & 547 & 1.86\dag\\
$1024^3$ & 113\ddag & 461 & 796 & 721 & 768 & 8.48\dag\\
$2048^3$ & 547 & 608 & 884 & 812 & 915 & 35.1\\
\bottomrule
\multicolumn{7}{l}{\dag TASK2 with broken sequential reduction.
  All values in GF/s.}\\
\multicolumn{7}{l}{\ddag OB at $1024^3$ (113\,GF/s) is anomalously low; see note below. All values in GF/s.}
\end{tabular}
\end{table}

\noindent The 6$\times$16 TASK1 kernel already achieves parity with OpenBLAS at 8T for medium-to-large sizes, and exceeds it at $N=128^3$ at 20T due to OpenBLAS thread-pool misconfiguration in the system build (see note below).\\

\textbf{Note on Table 2 vs.\ Table 3 OpenBLAS numbers.}
Table~\ref{tab:week2} reports OpenBLAS at $1024^3$, 20T as 113\,GF/s, while
Table~\ref{tab:week3} (same problem, same thread count) shows 481\,GF/s.
This $4.25\times$ discrepancy does \emph{not} reflect a fundamental performance
difference in OpenBLAS itself. The first measurements were collected during
an early benchmarking session (week 2 of the project) in which \texttt{OMP\_NUM\_THREADS} was set
per-process but the system OpenBLAS package was linked against a runtime that
capped its own thread pool at 8 threads by default. The next session (week 3 of the project) used
\texttt{OPENBLAS\_NUM\_THREADS=20} explicitly. All results in Table~\ref{tab:week3}
and beyond use this explicit setting and are directly comparable.

\subsection{Week~3 micro-kernel optimization}

\subsubsection{Loop unrolling $\times$4}
Extending the unroll factor from $\times$2 to $\times$4 increased arithmetic throughput by reducing loop overhead per FMA. This yielded 2--8\% speedup for medium and large matrix sizes but the AI agent needed explicit prompting to extend the unroll factor.

\subsubsection{Software prefetching}
Software prefetching uses explicit \texttt{\_mm\_prefetch} intrinsics to request that the CPU load data into L1 cache \emph{before} the compute instructions need it, overlapping memory latency with useful arithmetic and preventing pipeline stalls. Prefetch hints were placed within the 4-step unrolled loop to reduce L1 miss penalties. We load 2 cache lines of matrix $A$ and 4 cache lines of matrix $B$ per unrolled iteration.

\begin{lstlisting}[caption={6$\times$16 prefetch pattern inside the $\times$4 loop}]
_mm_prefetch((const char*)(pfA),      _MM_HINT_T0); // A CL0
_mm_prefetch((const char*)(pfA + 16), _MM_HINT_T0); // A CL1
pfA += 4 * MR;   // 96 bytes = 2 CLs ahead
_mm_prefetch((const char*)(pfB),      _MM_HINT_T0); // B CL0..CL3
_mm_prefetch((const char*)(pfB + 16), _MM_HINT_T0);
_mm_prefetch((const char*)(pfB + 32), _MM_HINT_T0);
_mm_prefetch((const char*)(pfB + 48), _MM_HINT_T0);
pfB += 4 * NR;   // 256 bytes = 4 CLs ahead
\end{lstlisting}

Software prefetching improved performance by 5--15\% for matrices between $512^3$ and $2048^3$.

\subsubsection{Non-temporal stores}
When the micro-kernel finishes computing a tile of $C$, that data will not be reused and could be written directly to memory. A normal store would first read the cache line into L3 (a \emph{read-for-ownership}), polluting it with data not needed and that evicts the still-needed $A$ and $B$ panels. Non-temporal stores (\texttt{\_mm256\_stream\_ps}) skip this read-for-ownership and write directly to memory through a write-combining buffer, freeing L3 capacity for the working set that actually benefits from caching. This requires 32-byte alignment, which we verify before applying the optimization:
\begin{lstlisting}
if (use_nt_store && ((uintptr_t)Cr % 32 == 0))
    _mm256_stream_ps(Cr, result);     // bypass L3 write
else
    _mm256_storeu_ps(Cr, result);
\end{lstlisting}
Non-temporal stores improve throughput by 5--20\% for large matrices ($2048^3$) but provide no benefit for small matrices where the L3 is underutilized.

\subsubsection{Assembly constraints}
When implementing the $6\times16$ micro-kernel in GCC extended inline assembly
(\texttt{asm volatile(...)}), we encountered the 30-operand limit: each C variable
passed into the assembly block counts as one operand, and the micro-kernel already
consumes nearly all 30 slots for pointers to $A$, $B$, $C$, loop strides, and
accumulator registers.
The loop trip count $k_4 = k\_\text{len}/4$ would have required one additional
operand, which exceeded the limit. This is a bug the AI agent ran into and struggled with for several minutes, but managed to solve on its own without needing human review of the code, only requesting human approval to try and implement the needed fixes.

The solution it finally came up with was to compute $k_4$ \emph{inside} the assembly block using two x86
instructions, avoiding the need for an extra operand:
\begin{lstlisting}[language={[x86masm]Assembler}]
"movl   %[klen], %%eax\n\t"   // copy k_len from C variable into register eax
"shrl   $2,      %%eax\n\t"   // logical right-shift by 2 == divide by 4
\end{lstlisting}

\noindent
In this code snippet:
\begin{itemize}
  \item \texttt{\%[klen]} is GCC's named-operand notation: the C variable
        \texttt{klen} is substituted here by the compiler.
  \item \texttt{\%\%eax} — the double percent sign is required inside GCC
        inline assembly to produce a literal \texttt{\%} before the register
        name; x86 general-purpose register \texttt{eax} holds a 32-bit integer.
  \item \texttt{movl} copies a 32-bit (\texttt{l} = \emph{long}) value from the
        source operand into \texttt{eax}.
  \item \texttt{shrl \$2} performs a logical right-shift by 2 bit positions,
        which is equivalent to an unsigned integer division by $2^2 = 4$ and is
        faster than an actual division instruction.
  \item \texttt{$\backslash$n$\backslash$t} at the end of each string literal is
        the standard way to separate assembly instructions inside a GCC
        \texttt{asm()} string: \texttt{$\backslash$n} starts a new line and
        \texttt{$\backslash$t} adds a tab so the disassembler output stays
        readable.
\end{itemize}

\subsubsection{Further structural fixes}

Three further limitations were addressed in the third week of the project. These are implemented using an autotuning loop:

\begin{itemize}
  \item \textbf{Small-matrix bypass}: Matrices with $MNK < 128^3$ are computed without tiling overhead. This improved $64^3$ performance from 34.1 to 42.3 GF/s (+24\%).
  \item \textbf{Adaptive MC parameter}: When $M/MC < 2 \cdot T$, MC is reduced so that $\lceil M/MC \rceil \ge 2T$, distributing M-tiles evenly across threads. At 20 threads on $128^3$, this raised performance from 75.3 to 126.6 GF/s (+68\%).
  % Diff 6: Adaptive NC from File 2
  \item \textbf{Adaptive NC parameter}: When $\lceil N/NC \rceil \cdot \lceil M/MC \rceil < T$ (too few 2-D tiles to saturate all threads), NC is reduced to $\lfloor N/(T \cdot NR) \rfloor \cdot NR$ so that the product of M- and N-tiles is at least $T$. Without this fix, $N=2048 < NC=4096$ produced only 18 tiles for 20 threads, triggering an unnecessary fallback to the less efficient 3-D K-replication mode. After the fix, tile count rises to $\sim$400, and 20T SGEMM at $N=2048$ improves by $+1.6\times$.
\end{itemize}

\subsubsection{Results at week 3 of the project}

% Diff 7: Table 3 structure from File 1 (1T/4T/8T)
\begin{table}[h]
\centering
\caption{SGEMM 6$\times$16 TASK1 at the end of week 3. Values in GF/s.}
\label{tab:week3}
\small
\begin{tabular}{lrrrrrrrr}
\toprule
& \multicolumn{2}{c}{1 thread} & \multicolumn{2}{c}{4 threads} & \multicolumn{3}{c}{8 threads}\\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}\cmidrule(lr){6-8}
Size & Ours & OB & Ours & OB & Ours & OB & Ratio\\
\midrule
$64^3$   &  50.7 &  50.5 &  50.8 &  50.7 &  50.7 &  50.8 & 1.00$\times$\\
$128^3$  &  62.7 &  62.7 & 108.2 & 106.6 &  99.2 &  98.4 & 1.01$\times$\\
$256^3$  &  67.9 &  68.2 & 200.6 & 200.3 & 290.3 & 292.6 & 0.99$\times$\\
$512^3$  &  70.1 &  71.2 & 228.3 & 229.3 & 409.3 & 406.3 & 1.01$\times$\\
$1024^3$ &  69.7 &  72.1 & 227.2 & 228.1 & 458.3 & 460.2 & 0.99$\times$\\
$2048^3$ &  69.9 &  69.6 & 247.1 & 247.6 & 505.7 & 505.5 & 1.00$\times$\\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Thread scaling.}
Performance scales linearly from 1T to 4T for all sizes $\ge 128^3$.
At 8T the implementation matches OpenBLAS within 1\% across all tested sizes,
confirming that the BLIS tiling strategy and adaptive MC/NC selection distribute
work evenly without incurring synchronisation overhead.

%%============================================================
\section{SGEMM parallel scheduling}
\label{sec:parallel}
%%============================================================

At this point in the project we have added multi-core parallelism to SGEMM. Next, we further compared two OpenMP scheduling models — loop-based and task-based — to understand their overhead characteristics. However, this was surprisingly difficult to implement using the AI agent. Indeed, from the start the agent was convinced that a task-based model would not perform well and thus, despite explicitly prompting it to use tasks, it continued using loops instead. Only when we explained forcefully and in great detail that a comparison of the two methods was needed to make any conclusions and implementation choices, did it finally produce a new model implementing both methods and allowing us to compare them.

\subsection{Loop-based parallelism}
Loop-based parallelism statically or dynamically partitions a nested \texttt{for} loop and assigns contiguous iteration ranges to threads before execution begins.
Because the work decomposition is determined at entry to the parallel region and the iteration space is a regular 2D grid, the OpenMP runtime overhead is $O(T)$ (one barrier per parallel region) and effectively zero per tile.

\subsection{Task-based parallelism}
Task-based parallelism treats each tile of $C$ as an independent work item. A master thread creates task descriptors — heap-allocated structs capturing all \texttt{firstprivate} variables — and pushes them into a runtime-managed queue; any idle thread dequeues and executes them.
This model supports irregular workloads and dependency graphs (via \texttt{depend} clauses), but incurs per-task allocation and scheduling overhead proportional to the number of tiles.

\subsection{Comparison and crossover}

We benchmarked the $6\times16$ micro-kernel using both paradigms in a 2D parallel configuration (TASK1) using 8 threads. 

\paragraph{Small matrices ($N \le 256$).}
For a $128 \times 128$ matrix, the work required to multiply a $6 \times 16$ tile is infinitesimal (measured in instructions). The time it takes OpenMP to allocate a block of memory, capture the local variables, push the task to a queue, and have a worker thread pop it is orders of magnitude larger than the math itself. \textbf{Task overhead completely dominates execution time.}

\paragraph{Large matrices ($N \ge 1024$).}
Matrix multiplication scales as $O(N^3)$ in arithmetic but only $O(N^2)$ in tile count. As $N$ grows, per-tile compute time dominates task-descriptor creation cost, and the two scheduling modes converge. By $N=2048$, the performance gap shrinks to $\sim$3.7\%.

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{semilogyaxis}[
  width=13cm, height=6.5cm,
  xtick={64,128,256,512,1024,2048},
  xticklabels={$64^3$,$128^3$,$256^3$,$512^3$,$1024^3$,$2048^3$},
  xlabel={Matrix size $N$},
  ylabel={GF/s (log scale)},
  legend style={at={(0.5,-0.28)},anchor=north,legend columns=2,font=\small},
  grid=both, grid style={dashed,gray!30},
  title={SGEMM Parallelism Comparison at 20 Threads},
  title style={font=\bfseries\small},
  ymin=1, ymax=1200,
]
\addplot[color=oborange,ultra thick,mark=diamond*,mark size=3]
  coordinates {(64,50.15)(128,56.88)(256,150.04)(512,216.54)(1024,481.34)(2048,532.88)};
\addlegendentry{OpenBLAS}

\addplot[color=ourblue,ultra thick,mark=square*,mark size=3]
  coordinates {(64,41.74)(128,104.54)(256,304.15)(512,536.68)(1024,751.89)(2048,692.80)};
\addlegendentry{TASK1 [\texttt{parallel for}] \textbf{$\leftarrow$ winner}}

\addplot[color=asmblue,thick,mark=triangle*,mark size=3,dashed]
  coordinates {(64,40.37)(128,102.45)(256,299.30)(512,572.27)(1024,766.56)(2048,741.36)};
\addlegendentry{ASM [\texttt{parallel for}]}

\addplot[color=taskred,thick,mark=*,mark size=3,dotted]
  coordinates {(64,41.82)(128,11.21)(256,26.28)(512,25.06)(1024,64.89)(2048,65.51)};
\addlegendentry{TASK1 [\texttt{taskloop}], 10$\times$ slower}

\addplot[color=task2green,thick,mark=x,mark size=4,densely dotted]
  coordinates {(64,41.82)(128,1.49)(256,3.23)(512,10.17)(1024,29.64)(2048,85.38)};
\addlegendentry{TASK2 [\texttt{parallel for}]}
\end{semilogyaxis}
\end{tikzpicture}
\caption{\texttt{taskloop} creates heap-allocated task objects for each tile. At $128^3$, tiles take $<$10$\,\mu$s to compute, resulting in a 10--14$\times$ overhead vs.\ \texttt{parallel for}. TASK2 (K-replication) is also slower due to private buffer reduction adding $r\!\times\!MC\!\times\!NC$ floats of memory traffic.}
\label{fig:parallel_modes}
\end{figure}

\paragraph{Dynamic scheduling.}
A \texttt{PARALLEL\_DYNAMIC} mode was added to the autotuning loop to select 2D or 3D automatically based on
$\lfloor M/MC\rfloor\cdot\lfloor N/NC\rfloor$ vs thread count. Despite working with an autotuning loop (explained in more detail in part 4) from as early as week 2, the AI agent never took initiative to update it with new tunable parameters and had to be explicitly prompted each time.

\paragraph{Residual \texttt{taskloop} in final code.}
During final benchmarking we observed that 8T SGEMM performance was indistinguishable
from 1T, a clear sign of serialization. Human review of \texttt{sgemm\_task1} revealed
that the computation loop had not been updated along with the packing loops: it still
used \texttt{\#pragma omp single\{taskloop collapse(2) grainsize(1)\}}, creating
$nr\_ic \times nr\_jt \approx 2304$ task descriptors per KC-step before any arithmetic
began. Replacing it with \texttt{\#pragma omp for collapse(2) schedule(static)}
eliminated the heap-allocation overhead and restored correct scaling
($+6.4\times$ at 8T for $N=2048$). The same \texttt{taskloop} pattern was present in
\texttt{sgemv} and \texttt{sger} (one task per row), causing degradation at 16T;
replacing with \texttt{parallel for} and a size threshold
($m \cdot n \ge 512 \cdot 1024$) corrected both kernels.

%%============================================================
\section{BLAS~3 derived kernels}
%%============================================================

BLAS 3 refers to a set of routines for matrix-matrix operations.

\subsection{SSYRK}

SSYRK performs a \textbf{symmetric rank-k update}: it computes $C = \alpha AA^T + \beta C$ .

\paragraph{First try.}
The AI agent delegated the computation to \texttt{sgemm\_ex(ib, jb, k, alpha, A\_i, lda,
\textbf{A\_j}, lda, ...)}. However, \texttt{sgemm\_ex} expects a
$K\times jb$ matrix as the second argument, yet $A_j$ is $jb\times K$.
As a result, all the tests failed (max error $\approx 0.4$ vs tolerance $10^{-3}$). The AI did not detect the argument layout mismatch at first but was able to debug without human intervention upon realizing that the tests had failed.

\paragraph{Second try.}
Each output element $C_{ii,jj} = \alpha \sum_k A_{ii,k}A_{jj,k}$ is computed
as an AVX2 dot product of two contiguous rows:
\begin{lstlisting}
const float *Ai = a + ii * lda, *Aj = a + jj * lda;
__m256 acc = _mm256_setzero_ps();
for (int kk = 0; kk <= k - 8; kk += 8)
    acc = _mm256_fmadd_ps(_mm256_loadu_ps(Ai+kk),
                          _mm256_loadu_ps(Aj+kk), acc);
// horizontal sum -> c[ii*ldc+jj] = alpha*s + beta*c[...]
\end{lstlisting}
% Diff 9: SSYRK throughput from File 1
With 128$\times$128 tiling for outer-loop L2 reuse and OMP dynamic parallel over triangular
tiles, all tests pass this time. Throughput is $\sim$9.5~GF/s ($n=256$, no panel
packing, deferred).

% Diff 10: STRSM and STRMM descriptions from File 1
\subsection{STRSM and STRMM}

\textbf{STRSM} solves a triangular matrix equation of the form $AX = \alpha B$, where A is a triangular matrix.
Our implementation uses \emph{panel-column blocking} (NB\,=\,32): rather than processing one column of $B$ at a time
(which causes strided memory access $b[k \cdot \text{ldb} + j]$), we solve NB columns simultaneously.
The inner update becomes a contiguous row-slice read — enabling auto-vectorisation by the compiler — and
the outer NB-wide panels are parallelised with \texttt{\#pragma omp parallel for}.\\\\
\textbf{STRMM} multiplies a general matrix $B$ by a triangular matrix $A$, overwriting $B$ in-place.
Our final implementation uses the same \emph{panel-column blocking} strategy as STRSM (NB\,=\,32 columns per panel):
threads are distributed over column panels of $B$; within each panel, rows are processed in the dependency order
(top-to-bottom for upper-triangular, bottom-to-top for lower).
This is the only correct in-place parallel approach — parallelising over rows instead creates a data race,
since row $i$ must read $B[k][\cdot]$ while thread $k$ concurrently writes $B[k][\cdot]$.\\

\textbf{Correctness challenge}: the test generator used uniform $U[0.5,1.5]$ diagonal in the triangular matrix,
producing condition numbers $\sim 3^n$ and relative errors of 5\%. It required a fix implemented by the AI agent:
\texttt{diagonal = 2.0 + rand()}, \texttt{off-diagonal = 0.01*rand()}.

\begin{table}[h]
\centering
\caption{BLAS~3 correctness and throughput (all tests vs.\ OpenBLAS reference).}
\small
\begin{tabular}{lll}
\toprule
Kernel & GF/s ($n=256$, 1T) & vs.\ OpenBLAS\\
\midrule
ssyrk (tiled dot product)    &  9.4 & 0.19$\times$ (no panel packing)\\
strsm (panel-blocked subst.) & 14.2 & 0.25$\times$ (NB=32, vectorised)\\
strmm (panel-col., in-place) &  5.9 & 0.19$\times$ (1T); best 0.34$\times$ ($N=128$, 4T)\\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Known limitations:}
We fixed the execution-layer bottlenecks (scheduling and tiling) for the core SGEMM, but reaching parity on derived kernels would require a second-order optimization: refactoring them to use the GEMM packing buffers directly instead of naive dot-products or dense copies. This could be a future optimization in this project:\\

\textbf{SSYRK.}
Our tiled dot-product approach computes $C_{ij} = \alpha\sum_k A_{ik}A_{jk}$
elementwise, with outer-loop tiling for L2 reuse. It avoids packing but
accesses $A$ column-wise for the second argument, causing cache-line
inefficiency for large $k$. OpenBLAS uses a GEMM-style panel packing
strategy: it copies panels of $A$ into a contiguous buffer ($MR\times KC$
or $NR\times KC$), effectively recasting SSYRK as a GEMM over the packed
panels. This eliminates the column-access penalty and achieves near-GEMM
arithmetic intensity. To match OpenBLAS, our SSYRK would need to implement
the same BLIS-style packing, extending the existing SGEMM infrastructure.\\

\textbf{STRSM.}
Panel blocking reduces the access penalty but does not eliminate the fundamental
$O(N^2)$ sequential dependency chain within each column panel: each row of $X$
requires all previous rows to be complete. OpenBLAS uses a right-looking blocked TRSM
that first solves a small diagonal block (fitting in registers), then updates the remaining
columns with a GEMM call — amortising the serial-dependency cost across large tile widths
and reaching near-GEMM memory traffic. Our panel approach closes part of this gap
(0.25$\times$ vs.\ 0.03$\times$ before), and the remaining factor of $\sim$4 would require
the full GEMM-update step.\\

\textbf{STRMM.}
Replacing the original row-parallel in-place approach (which had an undetected data race) with
panel-column blocking removed both the allocation overhead ($<$0.01$\times$ previously) and the
race condition.  The panel approach re-reads $A$ once per column panel ($n/\text{NB}$ passes),
which reduces single-thread efficiency compared to a serial row traversal, but enables correct
multithreaded execution: at $N=128$, 4T the implementation reaches 0.34$\times$ of OpenBLAS.
The remaining gap for large $N$ is due to the absence of panel packing: OpenBLAS fuses the
triangular structure into its $B$-panel packing routine, reaching near-GEMM arithmetic intensity.

%%============================================================
\section{Peak utilization}
%%============================================================

SGEMM is compute-bound (not bandwidth-bound) due to BLIS panel reuse. With
perfect tiling (A panel resident in L2, B panel in L3, C written via
non-temporal stores), each element of A is reused $NC/NR$ times and each
element of B is reused $MC/MR$ times. Dominant DRAM traffic reduces to one
pass of A and one pass of B:
\begin{equation}
  \mathrm{AI} = \frac{2N^3}{(N^2 + N^2)\times 4\;\mathrm{B/float}}
              = \frac{N}{4}\;\mathrm{FLOP/byte}
              \approx 256\;\mathrm{FLOP/byte} \quad (N{=}1024)
  \label{eq:sgemm_ai}
\end{equation}
This far exceeds the roofline crossover ($\approx 1.8$\,FLOP/byte at
73.6\,GF/s peak and 40\,GB/s measured DRAM bandwidth), confirming SGEMM is firmly in the
compute-bound regime for all practical matrix sizes.

% Diff 8: Peak utilization chart from File 1
\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=13cm, height=5.5cm,
  ybar=5pt, bar width=14pt,
  symbolic x coords={1T,4T,8T},
  xtick=data,
  xlabel={Thread count},
  ylabel={\% of $k$-core peak ($k\!\times\!73.6$\,GF/s)},
  legend style={at={(0.5,-0.35)},anchor=north,legend columns=2,font=\small},
  ymin=0, ymax=100,
  nodes near coords,
  nodes near coords style={font=\tiny,rotate=90,anchor=west},
  grid=major, grid style={dashed,gray!40},
  title={SGEMM $1024^3$ as a percentage of machine peak (73.6\,GF/s/core, no turbo)},
  title style={font=\bfseries\small},
]
\addplot[fill=ourblue!80]   coordinates {(1T,95)(4T,77)(8T,78)};
\addlegendentry{Ours}
\addplot[fill=oborange!80] coordinates {(1T,98)(4T,77)(8T,78)};
\addlegendentry{OpenBLAS}
\end{axis}
\end{tikzpicture}
\caption{\% of $k$-core peak for SGEMM $1024^3$ (baseline: 73.6\,GF/s per core at fixed 2.30\,GHz, no turbo).
  At 1T both implementations reach $\sim$95--98\% of single-core peak.
  At 4T and 8T both reach 77--78\% of $k$-core peak — limited by insufficient tile count
  at this matrix size (only $3\!\times\!1$ tiles with default MC=480, NC=4096).
  Table~\ref{tab:hw_peaks} lists the underlying measured GF/s values.}
\label{fig:machine_peak}
\end{figure}

%%============================================================
\section{AI-development audit}
%%============================================================
Overall in this first part of the project, using AI-driven development allowed us to be much faster and efficient in producing the code. From only reading the instructions it was able to produce fully functioning (though rarely fully optimized) code, and further prompting led it to fix any issues quite efficiently. The agent was particularly efficient at generating its own tests and benchmarks, but less so at reviewing and improving its own code unless it produced explicit bugs or failed tests. We had to review the code ourselves to catch some obvious issues (such as using a sequential reduction step instead of a binary tree). The auto-tuning script was also never updated to accommodate the progress of the project and we had to explicitly prompt the agent to do so each time. Surprisingly the agent could show stubbornness if we prompted it for a task it deemed not pertinent, such as trying a task-based model instead of loops. In this situation we had to argue with it several times for it to finally produce the requested code.
\begin{table}[h]
\centering
\caption{AI success/failure log for all SGEMM and BLAS~3 tasks.}
\label{tab:ai_audit}
\small
\begin{tabular}{p{5cm}p{2cm}p{6.5cm}}
\toprule
Task & Result & Notes\\
\midrule
8$\times$8 AVX2 micro-kernel    & \textbf{Success}  & Correct BLIS framework + FMA kernel.\\
6$\times$16, 4$\times$24 kernels & \textbf{Success} & All 16 YMM registers used correctly.\\
ASM kernel operand limit         & \textbf{\color{red}Failure} & Exceeded 30-op limit; manual shift-right fix.\\
TASK1 with \texttt{taskloop}     & \textbf{\color{red}Failure} & 10$\times$ overhead; caught by benchmark.\\
Recursive reduction tree (TASK2) & \textbf{Success}  & Correct OMP task dependencies.\\
Software prefetch (next-tile)    & \textbf{Success}  & Correct distances first try.\\
Non-temporal stores              & \textbf{Partial}  & Alignment guard omitted; caught in review.\\
Small-matrix bypass              & \textbf{Success}  & Correct threshold and VLA buffers.\\
Adaptive MC selection            & \textbf{Success}  & Correct formula; +68\% on $128^3$.\\
PARALLEL\_DYNAMIC dispatch       & \textbf{Success}  & Power-of-2 for binary reduction tree.\\
Gradient Descent optimizer       & \textbf{Success}  & Coordinate descent + restarts.\\
Bayesian Optimizer (GP+EI)       & \textbf{Success}  & GP class, RBF kernel, EI; scipy fallback.\\
Simulated Annealing              & \textbf{Success}  & Metropolis + cooling schedule.\\
SSYRK (first attempt)            & \textbf{\color{red}Failure} & Wrong argument layout to sgemm\_ex.\\
SSYRK (dot-product version)      & \textbf{Success}  & Correct AVX2 implementation.\\
STRSM (all 4 variants)           & \textbf{Success}  & Back-substitution for all cases.\\
STRMM                            & \textbf{Success}  & Buffer-copy approach correct.\\
Test generator (\texttt{rand\_tri}) & \textbf{\color{red}Failure} & Ill-conditioned; manual diagonal fix.\\
\midrule
\multicolumn{3}{l}{\textbf{4 failures / 18 tasks (22\% failure rate). Failures caught by tests, benchmarks or human code review.}}\\
\bottomrule
\end{tabular}
\end{table}

%%============================================================
\newpage
\part{BLAS~1}
%%============================================================
BLAS 1 refers to vector-vector operations, which operate on O(n) data and perform O(n) operations, offering little opportunity for cache reuse. We implemented 9 vector kernels: \texttt{sscal}, \texttt{scopy}, \texttt{sswap}, \texttt{saxpy}, \texttt{sdot}, \texttt{snrm2}, \texttt{sasum}, \texttt{isamax}, \texttt{srot}.

\begin{itemize}
    \item \texttt{sscal} --- scales a vector by a scalar: $x = \alpha x$
    \item \texttt{scopy} --- copies a vector $x$ into another vector $y$
    \item \texttt{sswap} --- swaps the contents of two vectors $x$ and $y$
    \item \texttt{saxpy} --- adds a scaled vector to another: $y = \alpha x + y$
    \item \texttt{sdot} --- computes the dot product of two vectors: $x^\top y$
    \item \texttt{snrm2} --- computes the Euclidean norm of a vector: $\|x\|_2$
    \item \texttt{sasum} --- computes the sum of absolute values: $\|x\|_1$
    \item \texttt{isamax} --- finds the index of the element with the largest absolute value
    \item \texttt{srot} --- applies a Givens rotation to a pair of vectors $(x, y)$
\end{itemize}

\section{BLAS~1 implementation strategy}

Unlike BLAS~3, BLAS~1 kernels are memory-bandwidth-bound, so the effect of vectorization and multi-threading depends on which bandwidth tier (L1/L2/L3/DRAM) the data occupies — this cannot be predicted without measurement.

We asked the AI to generate all 9 kernels in parallel, using a staged approach: scalar baseline, then AVX2 vectorization, then OpenMP parallelism. The working hypothesis was that by comparing each stage, we could identify where each technique contributes and where it causes overhead.

\paragraph{Implementation strategy.}

\begin{enumerate}
    \item \textbf{Classify by Operation Type}: Partition kernels into three classes:
    \begin{itemize}
        \item \textbf{Bandwidth-Limited} (sscal, scopy, sswap, saxpy): Perform $\sim$1 FLOP per byte; single thread saturates the \emph{single-core} L3 bandwidth, but multi-threading can exploit aggregate L3 bandwidth.
        \item \textbf{Latency-Limited Reductions} (sdot, snrm2, sasum): Accumulate results via long dependency chains; ALU stalls on FMA latency ($\sim$5 cycles).
        \item \textbf{Index-Based} (isamax): Search for maximum; data-dependent branching prevents vectorization.
    \end{itemize}
    \item \textbf{Bandwidth-Limited}: Use contiguous memory access ($\texttt{incx} = 1$) with AVX2 unrolling. For strided access ($\texttt{incx} > 1$), use gather intrinsics or scalar fallback. Parallelise with \texttt{omp parallel for} to exploit aggregate L3 bandwidth across cores.
    \item \textbf{Latency-Limited}: Use 4--8 independent accumulators (depends on register availability) to hide FMA latency. Parallelize with \texttt{\#pragma omp parallel for} after vectorization, because scalar ALU is idle waiting for results.
    \item \textbf{Index-Based}: Accept scalar implementation; vectorization not applicable for data-dependent branches like \texttt{isamax}.
    \item \textbf{Reduction Across Threads}: For multi-threaded reductions (e.g., parallel \texttt{sdot}), use atomic operations or work-stealing to minimize synchronization overhead.
\end{enumerate}

\paragraph{Results.}
The AI generated scalar and AVX2 variants on the first iteration. We measured the AVX2 impact: for reduction operations (\texttt{sdot}, \texttt{snrm2}, \texttt{sasum}), results showed 4.3$\times$ to 8.8$\times$ speedup. This gain was expected because the AI agent used multiple independent accumulators (\texttt{\_\_m256}) to allow FMA pipeline to fill during the long dependency chain. Interestingly, bandwidth-limited operations (\texttt{sscal}, \texttt{saxpy}) showed 0--1$\times$ speedup, suggesting that single-threaded core bandwidth was already saturated and wider SIMD did not help.\\

However, when we tested non-unit strides (\texttt{incx > 1}), the AI defaulted to scalar loops, eliminating vectorization gains. We hypothesized that \texttt{\_mm256\_i32gather\_ps} could restore SIMD, but benchmarking revealed gather itself was slow (serial on our CPU). Instead, we asked the AI to use gather only for very large strides, and unrolled scalar loops for modest strides (2--8). After guiding the AI with this constraint, performance recovered to 1.5--2$\times$ vs scalar.\\

When we asked the AI to parallelize reductions with \texttt{\#pragma omp} for 8 threads, the initial code used \texttt{\#pragma omp critical} on the accumulator. Benchmarking showed this caused severe lock contention. We noticed this while reviewing the code and reformulated the prompt: thread-local accumulators followed by a final reduction. This design achieved approximately 4$\times$ better scaling, nearly saturating 8-thread speedup.

\section{BLAS~1 performance analysis}

The kernels were developed in three stages — scalar, AVX2, multi-threaded — each benchmarked against OpenBLAS to isolate the contribution of each technique.

\subsection*{Stage 1 scalar baseline}
We began with a scalar baseline using simple \texttt{for} loops. This serves as a reference point: the upper bound on what a single core can achieve without architectural tricks. The measurements show wide variation: simple memory moves (\texttt{scopy}, \texttt{sscal}) reach 19--22\,GB/s (already near single-core L3 bandwidth), while reduction-heavy kernels (\texttt{sdot}, \texttt{snrm2}) stall at 1.7--1.7\,GFLOP/s. This pattern confirmed our hypothesis: some kernels hit the memory wall even at baseline, while others are bottlenecked by arithmetic pipeline depth.

% Diff 13: Table caption from File 1
\begin{table}[h]
\centering
\caption{Scalar Baseline vs OpenBLAS ($N = 1048576$, scalar 1T vs.\ OpenBLAS at default thread count during Stage~1 measurement; for controlled 1T comparison see Table~\ref{tab:blas1_peak}).}
\small
\begin{tabular}{llrrr}
\toprule
Kernel & Metric & Scalar Result & OpenBLAS Result & \% of OpenBLAS \\
\midrule
\texttt{sscal} & GB/s & 19.46 & 18.12 & 107\% \\
\texttt{scopy} & GB/s & 21.78 & 17.95 & 121\% \\
\texttt{saxpy} & GFLOP/s & 5.54 & 41.72 & \textbf{13\%} \\
\texttt{sdot} & GFLOP/s & 1.68 & 7.51 & \textbf{22\%} \\
\texttt{snrm2} & GFLOP/s & 1.70 & 5.55 & \textbf{30\%} \\
\bottomrule
\end{tabular}
\end{table}

Simple memory-move operations (\texttt{scopy}, \texttt{sscal}) reached 19--22\,GB/s — near single-core L3 bandwidth (the $N=10^6$ vector, 4\,MB, fits in L3; single-thread DRAM triad is only 7.8\,GB/s). Compute-involved kernels (\texttt{saxpy}, \texttt{sdot}) were substantially slower than OpenBLAS at this stage, motivating the vectorization and parallelism steps below.

\subsection*{Stage 2 SIMD vectorization}
We hypothesized that AVX2 would help reductions by hiding FMA latency through wider parallelism. Using four independent \texttt{\_\_m256} accumulators (processing 32 floats per loop iteration), we tested whether this overlapped FMA pipelines enough to improve throughput.

Results show a clear bifurcation: memory-bound kernels (\texttt{sscal}, \texttt{saxpy}) show 1.0--1.01$\times$ speedup, confirming the memory bus, not the ALU, is saturated. Reductions show 4.3--8.8$\times$ speedup from independent accumulators hiding the $\sim$5-cycle FMA latency. The two techniques address orthogonal bottlenecks, which suggested the next step: multi-threading might help bandwidth-bound kernels by letting multiple cores share L3 aggregate bandwidth.

\begin{lstlisting}[caption={AVX2 sdot inner loop with 4 independent accumulators to hide latency}]
__m256 sum0 = _mm256_setzero_ps();
__m256 sum1 = _mm256_setzero_ps();
__m256 sum2 = _mm256_setzero_ps();
__m256 sum3 = _mm256_setzero_ps();

for (int i = 0; i <= n - 32; i += 32) {
    sum0 = _mm256_fmadd_ps(_mm256_loadu_ps(&x[i]),    _mm256_loadu_ps(&y[i]),    sum0);
    sum1 = _mm256_fmadd_ps(_mm256_loadu_ps(&x[i+8]),  _mm256_loadu_ps(&y[i+8]),  sum1);
    sum2 = _mm256_fmadd_ps(_mm256_loadu_ps(&x[i+16]), _mm256_loadu_ps(&y[i+16]), sum2);
    sum3 = _mm256_fmadd_ps(_mm256_loadu_ps(&x[i+24]), _mm256_loadu_ps(&y[i+24]), sum3);
}
// Sum reduction follows...
\end{lstlisting}

\begin{table}[h]
\centering
\caption{AVX2 Impact over Scalar ($N = 1048576$).}
\small
\begin{tabular}{llrrr}
\toprule
Kernel & Metric & Scalar & AVX2 & Speedup over Scalar \\
\midrule
\texttt{sscal} & GB/s & 19.46 & 19.78 & 1.01$\times$ \\
\texttt{saxpy} & GFLOP/s & 5.54 & 5.56 & 1.00$\times$ \\
\texttt{sdot} & GFLOP/s & 1.68 & 7.32 & \textbf{4.35$\times$} \\
\texttt{snrm2} & GFLOP/s & 1.70 & 14.99 & \textbf{8.81$\times$} \\
\texttt{sasum} & GB/s & 3.39 & 30.01 & \textbf{8.85$\times$} \\
\bottomrule
\end{tabular}
\end{table}

\subsection*{Stage 3 multi-threading}
Since L3 is shared across all cores and its internal ring bandwidth far exceeds the DDR4 interface, multiple threads can jointly saturate it for cache-resident vectors. For $N=2^{20}$ (4\,MB per array, resident in L3), the results confirm this: \texttt{sscal} reaches 145\,GB/s at 8T ($4\times$ single-threaded), \texttt{sdot} reaches 20.6\,GFLOP/s at 8T ($2.7\times$). OpenBLAS stays at 37.7\,GB/s and 7.6\,GFLOP/s because it does not thread BLAS-1 for this vector size by default.

To interpret these numbers relative to hardware: peak per-core is $2 \text{ FMA/cycle} \times 2.30\,\text{GHz} = 73.6\,\text{GFLOP/s}$. Single-threaded \texttt{sdot} reaches 7.6\,GFLOP/s (10\% of peak), confirming the bottleneck is memory bandwidth, not compute. Multi-threaded \texttt{sdot} at 20.6\,GFLOP/s redistributes the pressure onto aggregate L3 bandwidth; perfect linear scaling (8T~$\to$~60\,GF/s) is prevented by the OMP thread-synchronisation barrier and L3 ring-bus contention when all 8 cores access the same 8\,MB working set simultaneously. At 20 threads, some kernels scale sublinearly due to NUMA effects, while others like \texttt{sscal} exceed the single-thread L3 ceiling by aggregating bandwidth across cores.

\paragraph{Hardware peak utilization.}
All five kernels with $N = 2^{20}$ (4\,MB per array) operate on data resident in L3. The relevant bandwidth ceiling at 8T is the measured L3 aggregate ($\approx 350$\,GB/s). For compute-reuse kernels the equivalent memory bandwidth is $\text{GF/s} \times \text{bytes/FLOP}$.

% Diff 13: Numbers from File 1
\begin{table}[h]
\centering
\caption{BLAS~1 hardware peak utilization (8T, $N=2^{20}$, data in 25\,MB L3).
  L3 agg.\ BW $\approx 350$\,GB/s; DRAM $\approx 40$\,GB/s (which is not the binding constraint here).}
\label{tab:blas1_peak}
\small
\begin{tabular}{lllrr}
\toprule
Kernel & Bottleneck & Our result (8T) & BW equiv.\ (GB/s) & \% L3 ceiling \\
\midrule
\texttt{sscal} & L3 BW (r+w) & 145\,GB/s   & 145 & 41\% \\
\texttt{scopy} & L3 BW (r+w) & 63\,GB/s    &  63 & 18\% \\
\texttt{saxpy} & L3 BW (r+rw)& 19.1\,GF/s  & 115$^\dagger$ & 33\% \\
\texttt{sdot}  & L3 BW (rr)  & 20.6\,GF/s  &  82$^\ddagger$& 23\% \\
\texttt{snrm2} & L3 BW (r)   & 31.7\,GF/s  &  63$^\S$      & 18\% \\
\bottomrule
\multicolumn{5}{l}{$^\dagger$ saxpy: 2\,F per 12\,B (read $x$, read+write $y$); $19.1 \times 6 = 115$\,GB/s.}\\
\multicolumn{5}{l}{$^\ddagger$ sdot: 2\,F per 8\,B (read $x$, read $y$); $20.6 \times 4 = 82$\,GB/s.}\\
\multicolumn{5}{l}{$^\S$ snrm2: 2\,F per 4\,B (read $x$ only); $31.7 \times 2 = 63$\,GB/s.}
\end{tabular}
\end{table}

All kernels are L3-bandwidth-bound, not compute-bound. Utilisation ranges from 18--41\% of the L3 ceiling at 8T. The gap from 100\% is explained by OMP thread-barrier and reduction overhead, cache-coherence traffic on the L3 ring bus when all 8 cores share the same working set, and the fact that \texttt{scopy}/\texttt{saxpy}/\texttt{sdot}/\texttt{snrm2} access 2--3 arrays simultaneously, increasing effective working-set pressure. \texttt{sscal} performs best (41\%) because it accesses a single array in-place with no coherence conflicts between cores.

\begin{table}[h]
\centering
\caption{Final Comparative Results ($N = 2^{20}$, 4\,MB per array, L3-resident). Our implementation: \textbf{8T OMP}.
  OpenBLAS: \textbf{1T} (BLAS-1 threading disabled by default in the system package).
  Ratios reflect thread-count difference (8T vs 1T), not algorithmic superiority.}
\label{tab:blas1_final}
\small
\begin{tabular}{llrrrrr}
\toprule
Kernel & Metric & Scalar & AVX2 & AVX2+OMP (8T) & OpenBLAS (1T) & Ratio \\
\midrule
\texttt{sscal} & GB/s     & 19.46 & 19.78 & \textbf{145.0}  & 37.7  & \textbf{3.8$\times$} \\
\texttt{scopy} & GB/s     & 21.78 & 22.10 & \textbf{63.1}   & 22.0  & \textbf{2.9$\times$} \\
\texttt{saxpy} & GFLOP/s  & 5.54  & 5.56  & \textbf{19.1}   & 5.5   & \textbf{3.5$\times$} \\
\texttt{sdot}  & GFLOP/s  & 1.68  & 7.32  & \textbf{20.6}   & 7.6   & \textbf{2.7$\times$} \\
\texttt{snrm2} & GFLOP/s  & 1.70  & 14.99 & \textbf{31.7}   & 5.85  & \textbf{5.4$\times$} \\
\bottomrule
\end{tabular}
\end{table}

\section{BLAS~1 AI development audit}
This part of the project was the easiest for the AI-agent to develop and it excelled at it, with very little human intervention needed. The main difficulty was getting results in-between the development stages to measure the impact of each. Indeed the AI agent would provide only final results rather than intermediate results after small changes. As we move forward in the project and more data is produced, the AI agent's understanding of our expectations improves and it better interprets and executes our prompts.
\begin{table}[h]
\centering
\caption{AI success/failure log for BLAS~1 (9 kernels, staged development).}
\label{tab:blas1_ai_audit}
\small
\begin{tabular}{p{4.5cm}p{2cm}p{7cm}}
\toprule
Task & Result & Notes \\
\midrule
Scalar baseline (all 9 kernels)
  & \textbf{Success}
  & Correct loops, correct stride handling on first attempt. \\
AVX2 with 4 independent accumulators (\texttt{sdot}, \texttt{snrm2}, \texttt{sasum})
  & \textbf{Success}
  & AI generated multiple-accumulator pattern unprompted; 4.3--8.8$\times$ speedup. \\
AVX2 bandwidth kernels (\texttt{sscal}, \texttt{saxpy}, \texttt{scopy})
  & \textbf{Partial}
  & Correct but 0--1$\times$ over scalar; AI did not predict bandwidth saturation. \\
Non-unit stride (\texttt{incx > 1}) via \texttt{\_mm256\_i32gather\_ps}
  & \textbf{\color{red}Failure}
  & Gather is serialised on this CPU; benchmarks showed regression. \\
Non-unit stride via unrolled scalar loops (human-guided fix)
  & \textbf{Success}
  & After explicit constraint, AI generated 1.5--2$\times$ improvement. \\
Multi-threaded reduction with \texttt{\#pragma omp critical}
  & \textbf{\color{red}Failure}
  & Severe lock contention; identified by benchmark (near-zero scaling). \\
Thread-local accumulators + final sequential reduction (human fix)
  & \textbf{Success}
  & $4\times$ better scaling; 8T speedup near ideal for large $N$. \\
\texttt{isamax} AVX2 vectorisation
  & \textbf{Partial}
  & AI attempted masked comparison; accepted scalar fallback after discussion. \\
\midrule
\multicolumn{3}{l}{\textbf{2 failures / 2 partial / 4 successes in 8 distinct tasks. All failures caught by benchmarks.}} \\
\bottomrule
\end{tabular}
\end{table}

%%============================================================
\newpage
\part{BLAS~2}
%%============================================================

% Diff 14: BLAS 2 intro from File 2
BLAS2 refers to matrix-vector operations, which operate on O(n²) data but perform O(n²) operations, offering limited cache reuse.

\section{BLAS~2 algorithms and architecture}

At arithmetic intensity $\approx 0.5$\,FLOP/byte (2 FLOPs per 8 bytes for \texttt{sgemv}), BLAS 2 kernels are memory-bound. Algorithmic restructuring — specifically loop fusion and access-pattern changes — is the primary lever for improving performance.

% Diff 15: ssyr2 framing from File 2
\paragraph{\texttt{ssyr2} fusion.}
The symmetric rank-2 update $A := \alpha x y^T + \alpha y x^T + A$ can be computed two ways. Initially the AI-agent implemented a naive approach calling rank-1 updates twice, reading and writing $A$ twice ($4 \times O(N^2)$ bandwidth). Upon catching this and reprompting it, we now instead fuse both updates into a single pass. For each row $i$, we load $\alpha x_i$ and $\alpha y_i$ into broadcast registers, then perform dual FMA operations on the loaded row of $A$. This cuts memory traffic to $2 \times O(N^2)$.

\begin{lstlisting}[caption={ssyr2 Fused AVX2 Inner Loop}]
__m256 v_x_i = _mm256_set1_ps(alpha * x[i]);
__m256 v_y_i = _mm256_set1_ps(alpha * y[i]);

for (int j = 0; j <= i - 8; j += 8) {
    __m256 v_A = _mm256_loadu_ps(&A[i * lda + j]);
    __m256 v_y = _mm256_loadu_ps(&y[j]);
    __m256 v_x = _mm256_loadu_ps(&x[j]);
    
    // Dual FMA fusion: A = A + (alpha*x_i)*y + (alpha*y_i)*x
    v_A = _mm256_fmadd_ps(v_x_i, v_y, v_A);
    v_A = _mm256_fmadd_ps(v_y_i, v_x, v_A);
    
    _mm256_storeu_ps(&A[i * lda + j], v_A);
}
\end{lstlisting}

As shown in Figure~\ref{fig:blas2_plots}, this change led to \textbf{1.2$\times$ to 2.2$\times$ speedup over OpenBLAS} across the tested matrix sizes.

\paragraph{\texttt{sger} limit.}
For the general rank-1 update \texttt{sger} ($A = \alpha x y^T + A$), the implementation uses row-streaming access. The memory pattern is prefetcher-friendly because rows are accessed linearly. At matrix sizes $N \ge 512$ (matrix $\le 16$\,MB, fitting within the 25\,MB L3), single-threaded performance reaches 38--39\,GB/s — the single-core L3 bandwidth ceiling, not DRAM (single-thread DRAM triad $\approx 7.8$\,GB/s). Both our implementation and OpenBLAS are L3-bound at this point, leaving no room for single-thread algorithmic improvement.

\paragraph{\texttt{sgemv} transpose.}
For the transpose case \texttt{sgemv} ($y = \alpha A^T x + \beta y$), the initial implementation from the AI agent used a column-major loop:
\begin{lstlisting}[caption={Naive Transpose Loop}]
for (int i = 0; i < n; i++)
    for (int j = 0; j < m; j++)
        y[i] += A[j * lda + i] * x[j];
\end{lstlisting}
Measurement showed 1.7\,GB/s (versus OpenBLAS's 32\,GB/s). C matrices are row-major, so incrementing \texttt{j} jumps by $lda$ bytes (4KB for $N=1024$). Every iteration fetches a new cache line and uses $\sim$4 bytes, wasting 60 bytes per line.

When we reprompted the AI to parallelize this directly, it applied \texttt{\#pragma omp} to the outer loop. This introduced a secondary problem: false sharing on output vector $y$—multiple threads writing adjacent elements in the same 64-byte line caused cache coherence overhead.

We proposed to rewrite the transposed form as a row-wise \texttt{saxpy}-like operation where each thread owns a slice of $y$ and iterates sequentially over rows. This eliminated both column-stride misses and false sharing but the AI needed human intervention to come up with this solution.

\paragraph{\texttt{ssymv} limit.}
\textbf{First prompt.} The AI generated a loop over the stored lower triangle with a conditional
\texttt{if (j <= i)} guard. The branching prevented auto-vectorisation; single-thread throughput
was 0.8\,GB/s.
\textbf{Second prompt.} The agent separated the diagonal, lower-triangle, and reflected (upper) accesses
into three independent loops, enabling AVX2 on the lower part. Performance improved to
1.6\,GB/s — still 14$\times$ below OpenBLAS (22\,GB/s). This can be explained by the fact that the reflected part accesses
column $j$ for all rows $i < j$, which strides by $\texttt{lda}$ bytes: each iteration fetches a new cache line for one float. This is the same column-stride issue as \texttt{sgemv} Transpose.
Fully resolving this requires a blocked panel layout: copying the symmetric matrix into a
row-major packed panel so that the reflected accesses become row-wise. This is another possible future improvement for the project.
OpenBLAS implements exactly this strategy (BLIS-style packing), achieving near-bandwidth
utilisation.

\section{BLAS~2 performance evaluation}

\begin{figure}[h]
\centering
\begin{minipage}{0.32\textwidth}
\begin{tikzpicture}
\begin{axis}[
  width=\textwidth, height=4.5cm,
  ybar, bar width=5pt,
  symbolic x coords={64,256,1024,4096},
  xtick=data,
  xlabel={$N$},
  ylabel={GB/s},
  ymin=0, ymax=45,
  grid=major, grid style={dashed,gray!30},
  title={\texttt{ssyr2} Algorithmic Win (1T)},
  title style={font=\bfseries\scriptsize},
  legend style={at={(0.5,-0.35)},anchor=north,legend columns=2,font=\tiny},
]
\addplot[fill=task2green] coordinates {(64,10.19) (256,29.56) (1024,27.94) (4096,13.38)};
\addplot[fill=oborange] coordinates {(64,5.55) (256,15.21) (1024,18.83) (4096,11.27)};
\legend{Ours, OB}
\end{axis}
\end{tikzpicture}
\end{minipage}\hfill
\begin{minipage}{0.32\textwidth}
\begin{tikzpicture}
\begin{axis}[
  width=\textwidth, height=4.5cm,
  ybar, bar width=5pt,
  symbolic x coords={1T,4T,8T,16T},
  xtick=data,
  xlabel={Threads},
  ylabel={GB/s},
  ymin=0, ymax=550,
  grid=major, grid style={dashed,gray!30},
  title={\texttt{sger} Scaling ($N=2048$)},
  title style={font=\bfseries\scriptsize},
  legend style={at={(0.5,-0.35)},anchor=north,legend columns=2,font=\tiny},
]
\addplot[fill=ourblue] coordinates {(1T,35.98) (4T,141.31) (8T,287.48) (16T,491.23)};
\addplot[fill=oborange] coordinates {(1T,36.49) (4T,141.96) (8T,284.80) (16T,498.91)};
\legend{Ours (fixed), OB}
\end{axis}
\end{tikzpicture}
\end{minipage}\hfill
\begin{minipage}{0.32\textwidth}
\begin{tikzpicture}
\begin{axis}[
  width=\textwidth, height=4.5cm,
  ybar, bar width=5pt,
  symbolic x coords={1T,4T,8T,16T},
  xtick=data,
  xlabel={Threads},
  ylabel={GB/s},
  ymin=0, ymax=110,
  grid=major, grid style={dashed,gray!30},
  title={\texttt{ssymv} Bottleneck ($N=1024$)},
  title style={font=\bfseries\scriptsize},
  legend style={at={(0.5,-0.35)},anchor=north,legend columns=2,font=\tiny},
]
\addplot[fill=taskred] coordinates {(1T,1.61) (4T,3.89) (8T,7.06) (16T,13.35)};
\addplot[fill=oborange] coordinates {(1T,22.79) (4T,71.71) (8T,93.09) (16T,92.43)};
\legend{Ours, OB}
\end{axis}
\end{tikzpicture}
\end{minipage}
\caption{BLAS 2 performance analysis. Left: \texttt{ssyr2} algorithmic fusion outpaces OpenBLAS significantly. Center: \texttt{sger} scaling after \texttt{taskloop}$\to$\texttt{parallel for} fix; our 16T result (491\,GB/s) matches OpenBLAS (499\,GB/s). Right: \texttt{ssymv} demonstrates a severe cache-miss bottleneck due to non-contiguous column access on the reflected triangle, contrasting with OpenBLAS's effective L2 panel blocking.}
\label{fig:blas2_plots}
\end{figure}

\paragraph{OpenBLAS comparison.}
We beat OpenBLAS in \texttt{ssyr2}, match it in \texttt{sger}, and outperform it in multi-threaded \texttt{sgemv}. OpenBLAS still outperforms us in \texttt{sgemv} Transpose and \texttt{ssymv}. The reason is cache-blocking: OpenBLAS uses assembly-level panel loops that fetch a block of matrix $A$ into the L2 cache, then compute multiple elements of the output simultaneously. Our row-wise implementation is correct but cannot exploit this reuse on transposed or symmetric accesses.

For \texttt{sger} at $N=2048$ and 16 threads, after replacing the per-row \texttt{taskloop} with \texttt{parallel for}, our implementation reaches 491\,GB/s, compared to OpenBLAS's 498\,GB/s — a gap of under 2\%. These values reflect \emph{L3 aggregate bandwidth} (the $N=2048$ matrix occupies $\sim$16\,MB, fitting within the 25\,MB L3), not DDR4; the figure is physically plausible given the L3 ring bandwidth on a 20-core Xeon.

\paragraph{Hardware peak utilization (BLAS~2).}
For BLAS~2 kernels, the relevant ceiling is L3 bandwidth for in-cache sizes. Table~\ref{tab:blas2_peak} shows single-thread utilization at $N=512$.

\begin{table}[h]
\centering
\caption{BLAS~2 single-thread bandwidth utilisation ($N=512$, data in L3).
  L3 single-core BW $\approx 39$\,GB/s (empirical: \texttt{sger} 1T peak at $N=512$ is 36\,GB/s, extrapolating to theoretical ceiling).}
\label{tab:blas2_peak}
\small
\begin{tabular}{llrr}
\toprule
Kernel & Result (1T) & L3 1T ceiling & \% peak \\
\midrule
\texttt{sger}   & 36\,GB/s  & 39\,GB/s & 92\% \\
\texttt{ssyr2}  & 30\,GB/s  & 39\,GB/s & 77\% \\
\texttt{ssyr}   & 20\,GB/s  & 39\,GB/s & 51\% \\
\texttt{sgemv}  (NoTrans) & 31\,GB/s  & 39\,GB/s & 79\% \\
\texttt{ssymv}  & 1.6\,GB/s & 39\,GB/s & 4\% \\
\bottomrule
\multicolumn{4}{l}{L3 single-core BW $\approx 39$\,GB/s (empirical ceiling; \texttt{sscal} 1T = 19.5\,GB/s already counts r+w).}
\end{tabular}
\end{table}

\noindent \texttt{ssymv} at 4\% peak confirms the column-stride cache pathology identified above; all other kernels operate at 51--92\% of the single-core L3 bandwidth ceiling.

\section{BLAS~2 AI development audit}

\begin{table}[h]
\centering
\caption{AI success/failure log for BLAS~2 (7 kernels).}
\label{tab:blas2_ai_audit}
\small
\begin{tabular}{p{4.5cm}p{2cm}p{7cm}}
\toprule
Task & Result & Notes \\
\midrule
\texttt{sgemv} NoTrans (row-wise loop)
  & \textbf{Success}
  & Correct AVX2 inner product per row; good prefetcher behaviour. \\
\texttt{sgemv} Transpose (naive column loop)
  & \textbf{\color{red}Failure}
  & 1.7\,GB/s vs.\ expected 32\,GB/s; column-stride miss pattern. \\
\texttt{sgemv} Transpose (row-wise saxpy reformulation)
  & \textbf{Success}
  & After human diagnosis, AI rewrote correctly; performance recovered. \\
\texttt{sgemv} Transpose parallelisation (naive outer loop)
  & \textbf{\color{red}Failure}
  & False sharing on output vector $y$ (adjacent elements in same cache line). \\
\texttt{sger} (row-streaming AVX2)
  & \textbf{Success}
  & Row-major access pattern; reached 38--39\,GB/s (DDR4 ceiling). Multi-thread: initial \texttt{taskloop grainsize(8)} caused 16T degradation (266\,GB/s); fixed with \texttt{parallel for} + size threshold, recovering 491\,GB/s (99\% of OpenBLAS). \\
\texttt{ssymv} (separated lower/reflected loops)
  & \textbf{Partial}
  & Correct, but reflected half retains column-stride misses. \\
\texttt{ssyr} (lower-triangle update)
  & \textbf{Success}
  & AVX2 per-row FMA; 40\% L3 utilisation. \\
\texttt{ssyr2} (naive: two rank-1 passes)
  & \textbf{Success}
  & Correct but 2$\times$ bandwidth overhead. \\
\texttt{ssyr2} fusion (dual FMA, single pass)
  & \textbf{Success}
  & AI proposed fusion when asked to minimise bandwidth; 2.2$\times$ win. \\
\texttt{strmv}, \texttt{strsv}
  & \textbf{Success}
  & Triangular access patterns generated correctly on first attempt. \\
\midrule
\multicolumn{3}{l}{\textbf{2 failures / 1 partial / 7 successes in 10 tasks. Failures caught by benchmarks and human review.}} \\
\bottomrule
\end{tabular}
\end{table}

%%============================================================
\newpage
% Diff 11: Part 4 title from File 2
\part{Testing, Optimization and Replicability}
%%============================================================

\section{Testing and numerical stability}

We developed an automated C testbench covering all BLAS~1, 2, and 3 kernels. Three aspects required particular care: matrix conditioning, floating-point associativity, and coverage of non-unit strides and all transpose variants.

\paragraph{Uniform random matrices.}
The most common mistake in testing linear algebra software is populating matrices with uniform random numbers, $U(-1, 1)$. For operations like triangular solvers (\texttt{strsm, strsv}), a uniformly populated matrix has a condition number $\kappa(A)$ that grows exponentially with the matrix dimension $N$. 
In floating-point arithmetic, the relative error of the output is bounded by $\epsilon \cdot \kappa(A)$, where $\epsilon$ is machine precision ($\approx 10^{-7}$ for FP32). For $N=512$, $\kappa(A)$ easily exceeds $10^6$, causing the solver to amplify tiny roundoff errors into massive macroscopic differences, causing completely valid code to fail the tests.

\paragraph{Matrix conditioning.}
To solve this, our testbench utilizes specialized data generators:
\begin{itemize}
    \item \textbf{Positive-Definiteness (\texttt{rand\_symm})}: For symmetric operations, we generate diagonally dominant symmetric matrices. We populate the matrix with $U(0,1)$, mirror it, and then add $N$ to the diagonal. This guarantees positive eigenvalues and absolute numerical stability.
    \item \textbf{Well-Conditioned Triangular (\texttt{rand\_tri})}: For solvers, we restrict off-diagonal elements to $0.01 \cdot U(0,1)$, while forcing the main diagonal to $2.0 + U(0,1)$. This restricts the condition number to $O(1)$, completely isolating algorithm flaws from mathematical artifacts.
\end{itemize}

\paragraph{Parallel reduction tolerance.}
Floating-point addition is not associative: $(a+b)+c \neq a+(b+c)$. Because our OpenMP implementations divide matrices across cores and accumulate partial sums in an unpredictable order, our results will legitimately differ from the sequential OpenBLAS reference.
To handle this, the testbench enforces a strict relative tolerance check:
$$ \frac{|x_{\text{ours}} - x_{\text{ref}}|}{\max(|x_{\text{ours}}|, |x_{\text{ref}}|) + 10^{-6}} \le \text{tol} $$
For standard operations, $\text{tol} = 10^{-3}$. For deep reduction operations like \texttt{ssyrk} where thousands of products are accumulated in parallel, the tolerance is safely relaxed to $5\times10^{-3}$. All 1287 BLAS 1 and 118 BLAS 2/3 test cases pass under these conditions.

%%============================================================
% Diff 12: Autotuning section structure from File 2
\section{Autotuning loop}
%%============================================================
Two Python scripts implement an autonomous compile-benchmark-tune cycle.
\texttt{sgemm\_tuner.py} implements the three black-box search algorithms (coordinate gradient descent, Bayesian optimisation, simulated
annealing), treating the SGEMM kernel as a function from block-size
parameters $(\text{MC}, \text{KC}, \text{NC})$ to measured GF/s.
\texttt{agent\_loop.py} wraps \texttt{sgemm\_tuner.py}: it writes a C header
with the chosen parameters, invokes \texttt{make}, runs the benchmark binary,
parses the output, and feeds the result back to the tuner.
The user specifies an evaluation budget; the loop terminates on convergence or budget exhaustion.
Three black-box algorithms were implemented to tune (MC, KC, NC), parallel mode (2D or 3D) and binary reduction tree depth (r), and compared
at fixed budget (25 evaluations) on SGEMM~$512^3$ (1~thread, 6$\times$16).

\begin{table}[h]
\centering
\caption{Autotuning results on SGEMM $512^3$ (1T, 6$\times$16). Default = 64.35~GF/s.}
\label{tab:autotune}
\small
\begin{tabular}{lrrlp{4.5cm}}
\toprule
Algorithm & Best GF/s & Evals & Wall time & Notes\\
\midrule
Coord.\ Gradient Descent & 66.82 & 34 & 2.7\,s& 3 restarts; finds optimum early\\
Bayesian Opt.\ (GP+EI)   & \textbf{66.92} & 25 & 4.0\,s & \textbf{Winner}; explores wider NC\\
Simulated Annealing      & 65.08 & 24 & 2.1\,s & Flat landscape limits SA\\
Random Search (baseline) & 64.95 &  8 & 0.7\,s & Competitive given only 8 evals\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=13cm, height=6.5cm,
  xlabel={Number of evaluations (\#kernel calls)},
  ylabel={Running best GF/s},
  legend style={at={(0.5,-0.28)},anchor=north,legend columns=2,font=\small},
  grid=major, grid style={dashed,gray!40},
  ymin=60, ymax=70,
  xmin=1, xmax=37,
  title={Autotuning convergence: SGEMM $512^3$, 1T, 6$\times$16 kernel},
  title style={font=\bfseries\small},
]
\addplot[color=ourblue,ultra thick,mark=square*,mark size=3]
  coordinates {
    (1,64.35)(2,64.35)(3,64.35)(4,64.35)(5,64.35)
    (6,65.20)(7,65.20)(8,65.20)(9,66.82)(10,66.82)
    (15,66.82)(20,66.82)(25,66.82)(30,66.82)(34,66.82)};
\addlegendentry{Gradient Descent (34 evals)}

\addplot[color=taskred,ultra thick,mark=*,mark size=3]
  coordinates {
    (1,64.35)(2,64.78)(3,64.78)(4,64.78)(5,64.78)
    (6,64.85)(7,64.85)(8,65.08)(9,65.08)(10,65.08)
    (15,65.08)(20,65.41)(21,65.41)(22,65.41)
    (23,66.92)(24,66.92)(25,66.92)};
\addlegendentry{Bayesian Opt.\ GP+EI (25 evals)}

\addplot[color=sagreen,ultra thick,mark=triangle*,mark size=3]
  coordinates {
    (1,64.35)(2,64.78)(3,64.85)(4,64.85)(5,64.85)
    (8,64.85)(12,64.85)(18,64.85)
    (19,65.08)(20,65.08)(24,65.08)};
\addlegendentry{Simulated Annealing (24 evals)}

\addplot[color=gray,thick,dashed,no marks]
  coordinates {(1,64.35)(35,64.35)};
\addlegendentry{Default config = 64.35\,GF/s}
\end{axis}
\end{tikzpicture}
\caption{Convergence curves (running maximum). All algorithms converge within
  4\% of each other because the performance landscape is flat: most block sizes
  keeping A in L2 and B in L3 give $\sim$64--65~GF/s. Gradient Descent reaches
  its best (66.82) at eval~9; BO reaches 66.92 at eval~23.}
\label{fig:autotuning}
\end{figure}

\paragraph{Tunable parameters.}
\begin{enumerate}
  \item \texttt{MC} — macro-block size along $M$ (targets L2 residency of the $A$ panel)
  \item \texttt{KC} — macro-block size along $K$ (targets L3 residency of the $B$ panel)
  \item \texttt{NC} — macro-block size along $N$ (controls tile count and thread balance)
  \item \texttt{parallel\_mode} — 2D tile parallelism vs.\ 3D K-replication
  \item \texttt{r\_tasks} — number of K-slices in K-replication mode
  \item Software Prefetch Distances: Tuning the \texttt{\_MM\_HINT\_T0} distances (0, 32, 64, 128 bytes) to fetch data into L1 cache exactly when the ALUs need it, compensating for DDR4 latency.
  \item Loop Unrolling Factors: Testing 1$\times$, 2$\times$, 4$\times$, and 8$\times$ inner-loop unrolling to balance instruction cache sizes against pipeline depth.
\end{enumerate}

\paragraph{Bayesian optimization.}
The optimizer builds a Gaussian Process surrogate model of the performance landscape and selects configurations by maximizing Expected Improvement (EI), balancing exploration of unknown parameters with exploitation of fast configurations.

The optimization discovered non-power-of-two block sizes (\texttt{MC=480, KC=256, NC=10432}) that are difficult to find by manual tuning, achieving 4\% speedup over mathematically derived GotoBLAS defaults.

\paragraph{Black-box limits.}
The agent loop also showed where parameter tuning is insufficient. For structural bottlenecks, such as $O(\mu s)$ task creation overhead for microsecond-scale \texttt{taskloop} work or column-stride misses in \texttt{ssymv}, changing prefetch distance or unroll factors has little effect. In practice, this creates a near-flat search landscape for methods such as Simulated Annealing.

\paragraph{Overall effect.}
The tuner recovered the final \textbf{10--15\% of hardware performance} over the manually chosen defaults (104\%--117\% of un-tuned baselines). The preceding 85\% of the gains came from algorithmic changes — loop restructuring, memory access pattern fixes, and scheduling primitive selection — which the black-box tuner cannot discover because they alter the code structure, not just numeric parameters.

\section{Agentic Optimization and Project Replicability}

An experimental component of the last two weeks was the development and evaluation of \textbf{Agentic Optimization}. This methodology shifts the focus from manual coding to the creation of high-level \textbf{Implementation Plans} that allow an AI agent to autonomously build, test, and optimize a library from scratch without any human intervention.

The goal was to evaluate if the performance reached through manual tuning (SIMD, tiling, multi-threading) could be replicated --- and sometimes exceeded --- by an automated system following a rigorous architectural blueprint.

\subsection{The ``Start-from-Scratch'' Workflow}
The agentic workflow is structured around three main pillars:

\paragraph{The Implementation Guide.}
Instead of writing the code directly, we prompted the AI to author comprehensive \texttt{IMPLEMENTATION\_GUIDE.md} documents. These guides serve as ``technical DNA,'' containing:
\begin{itemize}
    \item \textbf{Project Structure}: Standardized directory layouts for headers, source, and benchmarks.
    \item \textbf{Architectural Constraints}: Explicit requirements for AVX2 vectorization, software prefetching, and non-temporal stores.
    \item \textbf{Optimization Heuristics}: Guidance on loop unrolling factors and tiling strategies (e.g., $M_R \times N_R$ micro-kernel dimensions).
\end{itemize}

\subsection{Replicability \& AI-HPC}
While only a small part of our project was dedicated to this experiment, it showed promise that \textbf{HPC performance is reproducible}. By encoding expert knowledge into an implementation plan rather than a final codebase, we created a system that can adapt to different hardware environments. This shift suggests a future where HPC engineers cooperate with AI as \textbf{Architects of Plans}, while AI agents act as \textbf{Optimizers of Implementation}, ensuring that every cycle of the target hardware is utilized effectively without manual, error-prone micro-tuning.

%%============================================================
\newpage
\section{Results Summary}
%%============================================================

% Diff 17: Combined BLAS 1 (File 1 numbers + File 2 commentary) and BLAS 3 (both 1-8T parity and 20T advantage)
\paragraph{BLAS~1 on this workload.}
Our BLAS~1 implementation is faster than OpenBLAS at 8T, with up to \textbf{5.4$\times$} on \texttt{snrm2} ($N = 2^{20}$, 4\,MB, L3-resident). The ratio arises because OpenBLAS does not thread BLAS-1 operations by default (single-threaded reference), while our kernels use 8 OMP threads to aggregate L3 bandwidth. The comparison reflects a difference in design scope — OpenBLAS conserves thread overhead for small vectors; our library always threads above the 32\,K-element threshold. At 20 threads, some kernels scale sublinearly due to NUMA effects, while others like \texttt{sscal} exceed the single-thread L3 ceiling by aggregating bandwidth across cores.

\paragraph{BLAS~1 scaling.}
At single-thread, AVX2 contributes 4--9$\times$ speedup for reduction operations (latency hiding via multiple accumulators) and 0--1$\times$ for bandwidth-limited operations. Adding 8-thread OpenMP yields an additional 4--7$\times$ improvement for reduction kernels and 3--8$\times$ for bandwidth kernels (excluding \texttt{saxpy}, where OpenBLAS also threads at 8T, giving parity). Combined speedups over the scalar single-thread baseline reach 4--18$\times$ across all kernels, with \texttt{sscal} at 145\,GB/s (8T) and \texttt{snrm2} at 31.7\,GFLOP/s (8T) representing the bandwidth ceiling of L3 aggregate access.

\paragraph{BLAS~2 results.}
BLAS~2 made clear where algorithmic improvements are enough and where architecture dominates. In \texttt{ssyr2}, fused updates achieved up to 2.2$\times$ speedup (single-pass dual FMA instead of two rank-1 passes). After fixing the \texttt{taskloop} regression in \texttt{sger} and \texttt{sgemv}, multi-thread performance is competitive with OpenBLAS: \texttt{sger} reaches 491\,GB/s at 16T (vs OpenBLAS 499\,GB/s), and \texttt{sgemv} NoTrans scales to 6--9$\times$ speedup from 1T to 8T, exceeding OpenBLAS in this configuration. In contrast, \texttt{ssymv} (0.07$\times$ vs OpenBLAS) remains limited by access pattern: OpenBLAS uses panel blocking with better cache reuse, while our row-wise implementation still suffers from column-stride misses on the reflected triangle. Closing this gap would require a different blocked design and lower-level kernel work.

\paragraph{BLAS~3 results and tuning.}
At 1--8 threads, SGEMM matches OpenBLAS within 1\% across all tested sizes ($N = 64$--$2048$), confirming that the BLIS tiling strategy and adaptive MC/NC selection distribute work evenly without incurring synchronisation overhead. At single-thread we reach 95--96\% of single-core peak (73.6\,GFLOP/s), and at 8T we sustain 78\% of 8-core peak — comparable to OpenBLAS. At 20 threads, our implementation outperforms OpenBLAS by \textbf{1.3--2.5$\times$} for medium-to-large matrices ($N \ge 256$) due to better thread-count balancing via adaptive NC tile-count adjustment across the 20-core machine. The autotuning improvement over hand-tuned defaults is only 4\% (from 64.3\,GF/s to 66.92\,GF/s), indicating that the BLIS tiling strategy places the implementation already near the roofline limit. Bayesian optimization successfully identified non-obvious block sizes (\texttt{MC=480}), but the gains are strictly marginal because the fundamental loop structure is already sound.

\section{AI-assisted development}

\subsection*{What worked}

The BLIS 5-loop framework (correct loop nesting, packing buffers, pointer arithmetic)
came out right on the first attempt and compiled without modification.
Reaching the same point by reading the Goto and Van Zee papers alone
would have taken much longer; with the AI it took less than an afternoon.
The 6$\times$16 and 4$\times$24 micro-kernels were also correct on the first try,
including the register pressure analysis.

For reduction kernels (\texttt{sdot}, \texttt{snrm2}, \texttt{sasum}), the AI
independently used four independent accumulators without being asked.
This is the right technique for hiding 5-cycle FMA latency, and the AI had
apparently internalized it from context.

The \texttt{ssyr2} fusion was the most useful AI-generated insight.
When asked to minimise memory traffic, the AI proposed collapsing two rank-1
passes into one, halving the number of reads and writes to $A$.
The suggestion required understanding both the mathematical identity
$\alpha xy^T + \alpha yx^T = $ single loop and how to express it with dual FMAs.
The result was a 2.2$\times$ speedup over OpenBLAS.

The testbench also came out well: diagonally-dominant matrix construction, relative
tolerance for parallel reductions, clear error messages.
These are the parts of numerical testing that are easy to get wrong,
but that were very easy for the AI agent to implement.

\subsection*{Where it failed}

The recurring failure mode was that the AI could not predict hardware behavior
before seeing a benchmark.

\textit{Column-stride access in sgemv Transpose.}
The AI wrote the mathematically natural formulation:
iterate over columns of $A^T$, accumulate into $y$.
In a row-major layout, this strides by \texttt{lda} bytes per inner iteration,
fetching a new 64-byte cache line for every 4 bytes used.
Measured throughput: 1.7\,GB/s versus OpenBLAS's 32\,GB/s.
The AI's diagnosis was "try prefetching the next column,"
whereas the actual fix was rewriting the algorithm: iterate over rows of $A$
and accumulate partial contributions to all of $y$ via saxpy-style updates.
Once we described this, the AI implemented it correctly.
The AI was unable to make this diagnosis itself
from a performance number alone.

\textit{False sharing.}
After reformulating sgemv Transpose, we asked the AI to parallelise it with
\texttt{omp parallel for} over the outer loop.
Performance did not improve because threads were writing adjacent elements of $y$ in the same 64-byte cache
line, triggering coherence invalidation.
The fix was assigning contiguous chunks of $y$ to each thread
(private output ranges, no shared cache lines).
The AI implemented this once we identified the issue; it could not identify it on its own
from the symptoms.

\textit{Assembly operand limits.}
GCC's extended inline assembly supports at most 30 named operands.
The AI generated a 32-operand block for the 6$\times$16 kernel.
The fix was computing the loop counter inside the asm block with a shift
instruction rather than passing it as a named constraint.
This is not in any standard reference, and the AI did not know the limit
until the compiler reported it. However once the bug was apparent, the AI agent managed to correct it without needing our guidance.

\subsection*{What could have further improved AI assistance}

Two changes would have compressed the iteration cycles significantly.

First: direct benchmark access.
Each iteration took 30--60 seconds on our end (compile, run, read output,
summarise, prompt).
The AI could not observe the output directly.
For performance debugging, this usually meant describing symptoms in text rather
than sharing a hardware counter trace.

Second: hardware latency models.
The AI has no internalized model of DDR4 access latency ($\sim$60\,ns),
L3 ring bandwidth as a function of active core count, or NUMA first-touch
allocation policy.
Every time we said "this access pattern will miss caches on this machine,"
the AI could produce a correct fix.

\subsection*{Reflections}

Looking at the six weeks overall, the AI shortened the scaffolding phase
(framework generation, SIMD boilerplate, test infrastructure) massively.
For the optimization phase, it was more useful as an executor than a diagnostician:
we often had to identify ourselves what needed to change, and then the AI rewrote the code accordingly.

One of our concerns with this project was about passive learning.
The AI-generated code that actually worked taught us less than the code that failed, because we didn't have to study it thoroughly, and since we didn't write it ourselves we were much less familiar with it. Getting lost in our own project was a real concern, especially as it got more complex. It was very easy to overlook the parts that worked correctly and not try to understand them since we didn't need to fix them. This was definitely brought to our attention during the weekly meet-ups with our supervisor when we were sometimes unable to explain the behaviour or results of our code.
The taskloop overhead, column-stride pathology, and false sharing are things
we now understand better because we had to trace them by hand and had to make sure we understood the code and underlying concepts to be able to point out the issues.
In that sense the failures of the AI were actually the part
of the project with the highest educational density for us.

\section{Conclusion}

We built a working BLAS library over six weeks using AI-assisted development
and measured its performance against OpenBLAS across 19 kernels.
At 1--8 threads, SGEMM matches OpenBLAS within 1\% across all tested sizes,
confirming that the AI-generated BLIS framework reaches the same arithmetic intensity
and cache-reuse efficiency as a hand-optimised reference implementation.
At 20 threads, our implementation outperforms OpenBLAS by 1.3--2.5$\times$ for
medium-to-large matrices, largely because our adaptive tile-count balancing
distributes work more evenly across 20 cores than the system OpenBLAS build.
BLAS~1 kernels on large cache-resident vectors outperform single-threaded OpenBLAS
by large factors, though this comparison is not entirely fair since OpenBLAS
does not thread BLAS-1 by default.
BLAS~2 showed both an algorithmic win (\texttt{ssyr2}, 2.2$\times$ via fusion)
and a clear architectural ceiling (\texttt{ssymv}, 0.07$\times$ due to
column-stride cache misses we did not resolve).

One noticeable pattern across all three levels was that
the AI produced correct code quickly when the algorithm was well-specified,
but produced plausible-but-wrong code when the bottleneck was hardware-dependent.
Closing the gap between "plausible" and "fast" required measurement, diagnosis,
and reformulation, all done by hand.
The auto-tuning loop recovered the final 10--15\% once the algorithms were structurally sound.

\end{document}
